---
title: "Alevin"
---

```{r knitr, include = FALSE}
DOCNAME = "01B-alevin"
NOW <- Sys.time()

# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
    if (before) {
        print(paste("Start:", Sys.time()))
        NOW <<- Sys.time()
    } else {
        print(paste("Stop:", Sys.time()))
        print(Sys.time() - NOW)
    }
})

knitr::opts_chunk$set(
    autodep        = TRUE,
    cache          = TRUE,
    cache.path     = paste0("cache/", DOCNAME, "/"),
    cache.comments = FALSE,
    echo           = TRUE,
    error          = FALSE,
    fig.align      = "center",
    fig.width      = 10,
    fig.height     = 8,
    message        = FALSE,
    warning        = FALSE,
    timeit         = TRUE
)
```

```{r libaries, cache = FALSE}
# scRNA-seq
library("SingleCellExperiment")

# Plotting
library("UpSetR")
library("cowplot")
library("grid")

# Tidyverse
library("tidyverse")
```

```{r source, cache = FALSE}
source(here::here("R/load.R"))
source(here::here("R/annotate.R"))
source(here::here("R/output.R"))
```

```{r depends-paths}
sel_path <- here::here("data/processed/01-selected.Rds")

alevin_paths <- c(
    here::here("data/alevin/Org1"),
    here::here("data/alevin/Org2"),
    here::here("data/alevin/Org3")
)
```

```{r bpparam, cache = FALSE}
bpparam <- BiocParallel::MulticoreParam(workers = 10)
```

Introduction
============

The standard way to quanitfy 10x Chromium scRNA-seq data is using the Cell
Ranger platform that performs traditional alignment to a reference genome and
counts the reads overlapping annotated genes. An alternative appraoch is to
estimate expression levels directly against the transcriptome. We have done that
using an approach designed for scRNA-seq data called Alevin available in the
Salmon package. In this document we are going to compare the results of this
approach to what we get from Cell Ranger and our previous processing.

```{r load-sel, cache.extra = tools::md5sum(sel_path)}
if (file.exists(sel_path)) {
    selected <- read_rds(sel_path)
} else {
    stop("Selected dataset is missing. ",
         "Please run '01-preprocessing.Rmd' first.",
         call. = FALSE)
}

colData(selected)$BarcodeSample <- paste(
    colData(selected)$Barcode, 
    colData(selected)$Sample,
    sep = "-"
)
```

```{r load-alevin, cache.extra = tools::md5sum(alevin_paths)}
alevin <- readAlevin(alevin_paths, dataset = "Orgs123Alevin")
alevin <- annotateSCE(alevin, calc_qc = TRUE, BPPARAM = bpparam)

colData(alevin)$BarcodeSample <- paste(
    colData(alevin)$Barcode, 
    colData(alevin)$Sample,
    sep = "-"
)
```

```{r combine}
cell_data <- full_join(as.data.frame(colData(selected)),
                       as.data.frame(colData(alevin)),
                       by = "BarcodeSample",
                       suffix = c(".Trad", ".Alevin")) %>%
    mutate(Sample = Sample.Trad) %>%
    mutate(Sample = if_else(is.na(Sample), Sample.Alevin, Sample)) %>%
    select(BarcodeSample, Sample, contains("_"), contains("Filt"),
           -contains("control"), -contains("endogenous"), -contains("_MT")) %>%
    mutate_at(vars(contains("Filt")), replace_na, replace = FALSE) %>%
    mutate(AlevinFilt = !is.na(total_counts.Alevin)) %>%
    mutate(SelBy = "Both") %>%
    mutate(SelBy = if_else(!is.na(total_counts.Trad) &
                               is.na(total_counts.Alevin),
                           "Trad only", SelBy),
           SelBy = if_else(is.na(total_counts.Trad) &
                               !is.na(total_counts.Alevin),
                           "Alevin only", SelBy)) %>%
    mutate(SelBy = factor(SelBy,
                          levels = c("Both", "Trad only", "Alevin only")))

feat_data <- full_join(as.data.frame(rowData(selected)),
                       as.data.frame(rowData(alevin)),
                       by = "ID",
                       suffix = c(".Trad", ".Alevin")) %>%
    mutate(Annot = "Both") %>%
    mutate(Annot = if_else(!is.na(total_counts.Trad) &
                               is.na(total_counts.Alevin),
                           "Trad only", Annot),
           Annot = if_else(is.na(total_counts.Trad) &
                               !is.na(total_counts.Alevin),
                           "Alevin only", Annot)) %>%
    mutate(Annot = factor(Annot,
                          levels = c("Both", "Trad only", "Alevin only")))
```

Cell selection
==============

Alevin has it's own method of selecting cell-containing droplets. Let's see how
that compares to what we have done previously.

```{r cell-selection}
ggplot(cell_data, aes(x = SelBy, fill = Sample)) +
    geom_bar() +
    theme_minimal() +
    theme(axis.title.x = element_blank())
```

Cell counts {.tabset}
===========

Standard
--------

```{r cell-counts}
plot_data <- cell_data %>%
    mutate(Traditional = total_counts.Trad,
           Alevin = total_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("total_counts") +
    theme_minimal()
```

Logged
------

```{r cell-counts-log}
plot_data <- cell_data %>%
    mutate(Traditional = log10_total_counts.Trad,
           Alevin = log10_total_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("log10(total_counts)") +
    theme_minimal()
```

Cell features {.tabset}
=============

Standard
--------

```{r cell-features}
plot_data <- cell_data %>%
    mutate(Traditional = total_features_by_counts.Trad,
           Alevin = total_features_by_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("total_features_by_counts") +
    theme_minimal()
```

Logged
------

```{r cell-features-log}
plot_data <- cell_data %>%
    mutate(Traditional = log10_total_features_by_counts.Trad,
           Alevin = log10_total_features_by_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("log10(total_features_by_counts)") +
    theme_minimal()
```

Top 100 {.tabset}
=======

Percent counts in top 100 most expressed features.

```{r top-100}
plot_data <- cell_data %>%
    mutate(Traditional = pct_counts_in_top_100_features.Trad,
           Alevin = pct_counts_in_top_100_features.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("pct_counts_in_top_100_features") +
    theme_minimal()
```

Gene counts {.tabset}
===========

Standard
--------

```{r gene-counts}
plot_data <- feat_data %>%
    mutate(Traditional = total_counts.Trad,
           Alevin = total_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -1e5),
           Alevin = replace_na(Alevin, -1e5))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("total_counts") +
    theme_minimal()
```

Logged
------

```{r gene-counts-log}
plot_data <- feat_data %>%
    mutate(Traditional = log10_total_counts.Trad,
           Alevin = log10_total_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -1),
           Alevin = replace_na(Alevin, -1))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("log10(total_counts)") +
    theme_minimal()
```

Gene cells {.tabset}
==========

```{r gene-cells}
plot_data <- feat_data %>%
    mutate(Traditional = n_cells_by_counts.Trad,
           Alevin = n_cells_by_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -500),
           Alevin = replace_na(Alevin, -500))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("n_cells_by_counts") +
    theme_minimal()
```

Gene means {.tabset}
==========

Standard
--------

```{r gene-means}
plot_data <- feat_data %>%
    mutate(Traditional = mean_counts.Trad,
           Alevin = mean_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -5),
           Alevin = replace_na(Alevin, -5))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("mean_counts") +
    theme_minimal()
```

Logged
------

```{r gene-means-log}
plot_data <- feat_data %>%
    mutate(Traditional = log10_mean_counts.Trad,
           Alevin = log10_mean_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -0.1),
           Alevin = replace_na(Alevin, -0.1))

ggplot(plot_data, aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "blue") +
    ggtitle("log10(mean_counts)") +
    theme_minimal()
```

Counts-cells
============

```{r counts-zeros}
counts_data <- feat_data %>%
    filter(Annot == "Both") %>%
    select(ID, starts_with("log10_total_counts")) %>%
    gather(key = "Method", value = "log10_total_counts", -ID) %>%
    mutate(Method = str_split(Method, "\\.", simplify = TRUE)[, 2])

zeros_data <- feat_data %>%
    filter(Annot == "Both") %>%
    select(ID, starts_with("n_cells_by_counts")) %>%
    gather(key = "Method", value = "n_cells_by_counts", -ID) %>%
    mutate(Method = str_split(Method, "\\.", simplify = TRUE)[, 2])

plot_data <- left_join(counts_data, zeros_data, by = c("ID", "Method"))

ggplot(plot_data,
       aes(x = log10_total_counts, y = n_cells_by_counts, colour = Method)) +
    geom_point() +
    geom_smooth(aes(group = Method), colour = "blue") +
    facet_wrap(~ Method) +
    theme_minimal()
```

Figures
=======

```{r fig-plots, collapse = TRUE, fig.show = "hide"}
plot_data <- cell_data %>%
    select(BarcodeSample,
           `Cell Ranger v3` = CellRangerFilt,
           emptyDrops       = EmpDropsFilt,
           Alevin           = AlevinFilt) %>%
    mutate(`Cell Ranger v3` = if_else(`Cell Ranger v3`, 1L, 0L),
           emptyDrops       = if_else(emptyDrops,       1L, 0L),
           Alevin           = if_else(Alevin,           1L, 0L))

upset(plot_data, order.by = "freq",
      sets.x.label = "Number of cells",
      text.scale = c(1.5, 1.2, 1.5, 1, 1.5, 1.5),
      matrix.color = "#7A52C7",
      main.bar.color = "#7A52C7",
      sets.bar.color = "#7A52C7")
grid.edit("arrange", name = "arrange2")
comp_plot <- grid.grab()

plot_data <- cell_data %>%
    mutate(Traditional = total_counts.Trad,
           Alevin = total_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional,
                                    0.9 * min(Traditional, na.rm = TRUE)),
           Alevin = replace_na(Alevin, 0.9 * min(Alevin, na.rm = TRUE)))

counts_plot <- ggplot(plot_data,
                      aes(x = Traditional, y = Alevin, colour = SelBy)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "#7A52C7") +
    geom_smooth(data = filter(plot_data, SelBy == "Both"),
                method = "lm", colour = "#00B7C6") +
    scale_x_log10(labels = scales::number) +
    scale_y_log10(labels = scales::number) +
    scale_colour_manual(values = c("#EC008C", "#00ADEF", "#8DC63F"),
                        labels = c("Both", "Alignment based only",
                                   "Alevin only")) +
    ggtitle("Total counts per cell") +
    xlab("Alignment based") +
    theme_minimal() +
    theme(legend.position = "bottom",
          legend.title = element_blank())

plot_data <- feat_data %>%
    mutate(Traditional = log10_mean_counts.Trad,
           Alevin = log10_mean_counts.Alevin) %>%
    mutate(Traditional = replace_na(Traditional, -0.1),
           Alevin = replace_na(Alevin, -0.1))

mean_plot <- ggplot(plot_data,
                    aes(x = Traditional, y = Alevin, colour = Annot)) +
    geom_point(alpha = 0.3) +
    geom_abline(intercept = 0, slope = 1, colour = "#7A52C7") +
    geom_smooth(data = filter(plot_data, Annot == "Both"),
                method = "lm", colour = "#00B7C6") +
    scale_colour_manual(values = c("#EC008C", "#00ADEF", "#8DC63F"),
                        labels = c("Both", "Alignment based only",
                                   "Alevin only")) +
    ggtitle("Mean counts per gene (log10)") +
    xlab("Alignment based") +
    theme_minimal() +
    theme(legend.position = "bottom",
          legend.title = element_blank())

counts_data <- feat_data %>%
    filter(Annot == "Both") %>%
    select(ID, starts_with("log10_total_counts")) %>%
    gather(key = "Method", value = "log10_total_counts", -ID) %>%
    mutate(Method = str_split(Method, "\\.", simplify = TRUE)[, 2])

zeros_data <- feat_data %>%
    filter(Annot == "Both") %>%
    select(ID, starts_with("n_cells_by_counts")) %>%
    gather(key = "Method", value = "n_cells_by_counts", -ID) %>%
    mutate(Method = str_split(Method, "\\.", simplify = TRUE)[, 2]) %>%
    mutate(PropZero = if_else(Method == "Trad",
                              1 - n_cells_by_counts / ncol(selected),
                              1 - n_cells_by_counts / ncol(alevin)))

plot_data <- left_join(counts_data, zeros_data, by = c("ID", "Method")) %>%
    mutate(Method = factor(Method, levels = c("Trad", "Alevin"),
                           labels = c("Alignment based", "Alevin")))

zeros_plot <- ggplot(plot_data,
       aes(x = log10_total_counts, y = PropZero, colour = Method)) +
    geom_point(alpha = 0.3) +
    geom_smooth(aes(group = Method), colour = "#EC008C") +
    scale_y_continuous(labels = scales::percent) +
    scale_colour_manual(values = c("#00ADEF", "#8DC63F")) +
    facet_wrap(~ Method) +
    ggtitle("Zeros relationship") +
    xlab("Total counts per gene (log10)") +
    ylab("Percentage zeros") +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r make-fig}
fig <- plot_grid(comp_plot, counts_plot, mean_plot, zeros_plot,
          nrow = 2, labels = "AUTO")

ggsave(here::here("output", DOCNAME, "alevin-comparison.pdf"), fig,
       width = 7, height = 6, scale = 1.5)
ggsave(here::here("output", DOCNAME, "alevin-comparison.png"), fig,
       width = 7, height = 6, scale = 1.5)

fig
```


Summary
=======

Overall the two approaches seem to produce similar results. There are
differences between them but it is difficult to tell if one approach is
inaccurate. For the rest of this analysis we will stick with the traditional
approach as it is more familiar.

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    list(
        Parameter = "n_cells",
        Value = ncol(alevin),
        Description = "Number of cells in Alevin dataset"
    ),
    list(
        Parameter = "n_genes",
        Value = nrow(alevin),
        Description = "Number of genes in Alevin dataset"
    )
)

params <- jsonlite::toJSON(params, pretty = TRUE)
knitr::kable(jsonlite::fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r save}
write_rds(alevin, here::here("data/processed/01B-alevin.Rds"))
```

```{r output}
dir.create(here::here("output", DOCNAME), showWarnings = FALSE)

readr::write_lines(params, here::here("output", DOCNAME, "parameters.json"))

knitr::kable(data.frame(
    File = c(
        getDownloadLink("parameters.json", DOCNAME),
        getDownloadLink("alevin-comparison.png", DOCNAME),
        getDownloadLink("alevin-comparison.pdf", DOCNAME)
    ),
    Description = c(
        "Parameters set and used in this analysis",
        "Alevin comparison figure (PNG)",
        "Alevin comparison figure (PDF)"
    )
))
```

Session information
-------------------

```{r session-info, cache = FALSE}
devtools::session_info()
```
