---
title: "Quality control"
---

```{r knitr, include = FALSE}
DOCNAME = "02-quality-control"
NOW <- Sys.time()

# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
    if (before) {
        print(paste("Start:", Sys.time()))
        NOW <<- Sys.time()
    } else {
        print(paste("Stop:", Sys.time()))
        print(Sys.time() - NOW)
    }
})

knitr::opts_chunk$set(
    autodep        = TRUE,
    cache          = TRUE,
    cache.path     = paste0("cache/", DOCNAME, "/"),
    cache.comments = FALSE,
    echo           = TRUE,
    error          = FALSE,
    fig.align      = "center",
    fig.width      = 10,
    fig.height     = 8,
    message        = FALSE,
    warning        = FALSE,
    timeit         = TRUE
)
```

```{r libaries, cache = FALSE}
# scRNA-seq
library("SingleCellExperiment")
library("scater")
library("scran")

# RNA-seq
library("edgeR")

# Plotting
library("cowplot")
library("gridExtra")

# Tidyverse
library("tidyverse")
```

```{r source, cache = FALSE}
source(here::here("R/plotting.R"))
source(here::here("R/output.R"))
```

```{r depends-paths}
sel_path <- here::here("data/processed/01-selected.Rds")
```

```{r bpparam, cache = FALSE}
bpparam <- BiocParallel::MulticoreParam(workers = 10)
```

Introduction
============

In this document we are going to explore the dataset and filter it to remove
low quality cells and lowly expressed genes using the `scater` and `scran`
packages. The goal is to have a high quality dataset that can be used for
further analysis. 

```{r load, cache.extra = tools::md5sum(sel_path)}
if (file.exists(sel_path)) {
    selected <- read_rds(sel_path)
} else {
    stop("Selected dataset is missing. ",
         "Please run '01-preprocessing.Rmd' first.",
         call. = FALSE)
}

set.seed(1)
sizeFactors(selected) <- librarySizeFactors(selected)
selected <- normalize(selected)
selected <- runPCA(selected, method = "irlba")
selected <- runTSNE(selected)
selected <- runUMAP(selected)

cell_data <- as.data.frame(colData(selected))
feat_data <- as.data.frame(rowData(selected))
```

Exploration
===========

We will start off by making some plots to explore the dataset.

Expression by cell {.tabset}
------------------

Distributions by cell. Blue line shows the median and red lines show median
absolute deviations (MADs) from the median.

### Total counts {.unnumbered}

```{r counts}
outlierHistogram(cell_data, "log10_total_counts", mads = 1:6)
```

### Total features {.unnumbered}

```{r features}
outlierHistogram(cell_data, "log10_total_features_by_counts", mads = 1:6)
```

### Percent mitochondrial {.unnumbered}

```{r percent-MT}
outlierHistogram(cell_data, "pct_counts_MT", mads = c(1, 3, 5, 7))
```

Dimensionality reduction {.tabset}
------------------------

Dimensionality reduction plots coloured by technical factors can help identify
which may be playing a bit role in the dataset.

```{r dimred-factors}
dimred_factors <- c(
    "Sample"              = "Sample",
    "Total counts"        = "log10_total_counts",
    "CellCycle"           = "CellCycle", 
    "Total features"      = "log10_total_features_by_counts",
    "Mitochondrial genes" = "pct_counts_MT",
    "Selection method"    = "SelMethod"
)
```

### PCA

```{r pca}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotPCA(selected, colour_by = dimred_factors[fct_name],
            shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### t-SNE

```{r t-sne}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotTSNE(selected, colour_by = dimred_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### UMAP

```{r umap}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotUMAP(selected, colour_by = dimred_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

Explanatory variables
---------------------

This plot shows the percentage of variance in the dataset that is explained by
various technical factors. 

```{r exp-vars}
exp_vars <- c("Sample", "CellCycle", "SelMethod", "log10_total_counts",
              "pct_counts_in_top_100_features", "total_features_by_counts",
              "pct_counts_MT")

all_zero <- Matrix::rowSums(counts(selected)) == 0
plotExplanatoryVariables(selected[!all_zero, ], variables = exp_vars)
```

Here we see that factors associated with the number of counts and features in
each cell represent large sources of variation.

Cell cycle
----------

The dataset has already been scored for cell cycle activity. This plot shows
the G2/M score against the G1 score for each cell and let's us see the balance
of cell cycle phases in the dataset.

```{r cell-cycle}
ggplot(cell_data,
       aes(x = G1Score, y = G2MScore, colour = CellCycle, shape = Sample)) +
    geom_point() +
    xlab("G1 score") +
    ylab("G2/M score") +
    theme_minimal()

knitr::kable(table(Phase = cell_data$CellCycle, useNA = "ifany"))
```

Expression by gene {.tabset}
------------------

Distributions by cell. Blue line shows the median and red lines show median
absolute deviations (MADs) from the median. We show distributions for all genes
and those that have at least one count.

### Mean

```{r exprs-mean}
outlierHistogram(feat_data, "log10_mean_counts", mads = c(10, 20))
```

### Total

```{r exprs-total}
outlierHistogram(feat_data, "log10_total_counts", mads = 1:5)
```

### Mean (expressed)

```{r exprs-mean-expressed}
outlierHistogram(feat_data[feat_data$total_counts > 0, ],
                 "log10_mean_counts", mads = c(10, 20))
```

### Total (expressed)

```{r exprs-total-expressed}
outlierHistogram(feat_data[feat_data$total_counts > 0, ],
                 "log10_total_counts", mads = 1:5)
```

High expression genes
---------------------

We can also look at the expression levels of just the top 50 most expressed
genes.

```{r high-exprs}
plotHighestExprs(selected)
```

As is typical for 10x experiments we see that many of these are ribosomal
proteins.

Expression frequency {.tabset}
--------------------

The relationshop between the number of cells that express a gene and the overall
expression level can also be interesting. We expect to see that higher expressed
genes are expressed in more cells but there will also be some that stand out
from this.

### Frequency by mean

```{r freq-by-mean}
plotExprsFreqVsMean(selected, controls = NULL)
```

### Zeros by total counts

```{r ncells-by-mean}
ggplot(feat_data,
       aes(x = log10_total_counts, y = 1 - n_cells_by_counts / nrow(selected),
           colour = gene_biotype)) +
    geom_point(alpha = 0.2, size = 1) +
    scale_y_continuous(labels = scales::percent) +
    xlab("log(total counts)") +
    ylab("Percentage zeros")
```

Cell filtering
==============

We will now perform filtering to select high quality cells. Before we start we
have `r ncol(selected)` cells.

Manual outliers
---------------

The simplest filtering method is to set thresholds on some of the factors we
have explored. Specifically these are the total number of counts per cell, the
number of features expressed in each cell and the percentage of counts assigned
to genes on the mitochondrial chromosome which is used as a proxy for cell
damage. The selected thresholds and numbers of filtered cells using this method
are:

```{r manual-outliers}
counts_mads   <- 4
features_mads <- 4
mt_mads       <- 4

counts_out   <- isOutlier(cell_data$log10_total_counts,
                          nmads = counts_mads, type = "lower")
features_out <- isOutlier(cell_data$log10_total_features_by_counts,
                          nmads = features_mads, type = "lower")
mt_out       <- isOutlier(cell_data$pct_counts_MT,
                          nmads = mt_mads, type = "higher")

counts_thresh   <- attr(counts_out, "thresholds")["lower"]
features_thresh <- attr(features_out, "thresholds")["lower"]
mt_thresh       <- attr(mt_out, "thresholds")["higher"]

kept_manual <- !(counts_out | features_out | mt_out)
colData(selected)$KeptManual <- kept_manual

manual_summ <- tibble(
    Type = c(
        "Total counts",
        "Total features",
        "Mitochondrial %",
        "Kept (manual)"
    ),
    Threshold = c(
        paste("< 10 ^", round(counts_thresh, 2)),
        paste("< 10 ^", round(features_thresh, 2)),
        paste(">", round(mt_thresh, 2)),
        ""
    ),
    Count = c(
        sum(counts_out),
        sum(features_out),
        sum(mt_out),
        sum(kept_manual)
    )
)

knitr::kable(manual_summ)
```

PCA outliers
------------

An alternative approach is to perform a PCA using technical factors instead of
gene expression and then use outlier detection to identify low-quality cells.
This has the advantage of being an automated process but is difficult to
interpret and may be affected if there are many low-quality cells.

```{r pca-outliers}
selected <- runPCA(selected, use_coldata = TRUE, detect_outliers = TRUE)

plotReducedDim(selected, use_dimred = "PCA_coldata", colour_by = "outlier",
               add_ticks = FALSE)

kept_pca <- !colData(selected)$outlier
colData(selected)$KeptPCA <- kept_pca

pca_summ <- tibble(
    Type = c(
        "PCA",
        "Kept (PCA)"
    ),
    Count = c(
        sum(!kept_pca),
        sum(kept_pca)
    )
)

knitr::kable(pca_summ)
```

Outlier summary {.tabset}
---------------

```{r outlier-summary}
kept_both <- kept_manual & kept_pca
colData(selected)$Kept              <- "False"
colData(selected)$Kept[kept_manual] <- "Manual"
colData(selected)$Kept[kept_pca]    <- "PCA"
colData(selected)$Kept[kept_both]   <- "Both"
```

The manual method has identified `r sum(kept_manual)` high-quality cells and the
PCA method identified `r sum(kept_pca)`. If we combine both methods we get
`r sum(kept_both)` cells. We can look at which cells each method identified on
our dimensionality reduction plots.

```{r out-factors}
out_factors <- c(
    "Kept"                = "Kept",
    "Total counts"        = "log10_total_counts",
    "Total features"      = "log10_total_features_by_counts",
    "Mitochondrial genes" = "pct_counts_MT"
)
```

### PCA

```{r pca-summary}
plot_list <- lapply(names(out_factors), function(fct_name) {
    plotPCA(selected, colour_by = out_factors[fct_name],
            shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### t-SNE

```{r tSNE-summary}
plot_list <- lapply(names(out_factors), function(fct_name) {
    plotTSNE(selected, colour_by = out_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### UMAP

```{r umap-summary}
plot_list <- lapply(names(out_factors), function(fct_name) {
    plotUMAP(selected, colour_by = out_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

Filter
------

We are going to use the manual filtering as it is easier to interpret and
adjust.

```{r filter}
filtered <- selected[, kept_manual]

set.seed(1)
sizeFactors(filtered) <- librarySizeFactors(filtered)
filtered <- normalize(filtered)
filtered <- runPCA(filtered, method = "irlba")
filtered <- runTSNE(filtered)
filtered <- runUMAP(filtered)
```

Our filtered dataset now has `r ncol(filtered)` cells.

Doublets
========

One kind of low-quality cell we have not yet considered are doubletes where two
cells are captured in the same droplet. We expect doublets to have at least the
normal amount of RNA so they won't be captured by our previous filters. To
identify them we will use a method that simulates doublets by combining cells
and then checks the neighbourhood of cells to see if they are near simulated
doublets. This gives us a doublet score for each cell.

```{r doublets}
set.seed(1)
doublet_scores <- doubletCells(filtered, approximate = TRUE, BPPARAM = bpparam)

colData(filtered)$DoubletScore <- doublet_scores
```

Plots {.tabset}
-----

### PCA

```{r pca-doublets}
plotPCA(filtered, colour_by = "DoubletScore", add_ticks = FALSE)
```

### t-SNE

```{r tSNE-doublets}
plotTSNE(filtered, colour_by = "DoubletScore", add_ticks = FALSE)
```

### UMAP

```{r umap-doublets}
plotUMAP(filtered, colour_by = "DoubletScore", add_ticks = FALSE)
```

### Histogram

```{r hist-doublets}
outlierHistogram(as.data.frame(colData(filtered)), "DoubletScore",
                 mads = c(1, 3, 5, 7, 9, 11, 13, 15))
```

Filter
------

```{r filter-doublets}
doublet_out <- isOutlier(doublet_scores, nmads = 15, type = "higher")
doublet_thresh <- attr(doublet_out, "thresholds")["higher"]

filtered <- filtered[, !doublet_out]
```

We choose a threshold for the doublet score that removes around the number of
cells as the expected doublet frequency for a 10x experiment (around 1 percent).
A threshold of `r round(doublet_thresh, 2)` identifies `r sum(doublet_out)`
doublets leaving us with `r ncol(filtered)` cells.

Gene filtering
==============

We also want to perform som filtering of features to remove lowly expressed
genes that increase the computation required and may not meet the assumptions
of some methods. Let's look as some distributions now that we have removed
low-quality cells.

Distributions {.tabset}
-------------

### Average counts

```{r mean-histogram}
avg_counts <- calcAverage(filtered, use_size_factors = FALSE)
rowData(filtered)$AvgCount <- avg_counts
rowData(filtered)$Log10AvgCount <- log10(avg_counts)

outlierHistogram(as.data.frame(rowData(filtered)), "Log10AvgCount",
                 mads = 1:3, bins = 100)
```

### Number of cells

```{r cells-histogram}
outlierHistogram(as.data.frame(rowData(filtered)), "n_cells_by_counts",
                 mads = 1:3, bins = 100) +
    scale_x_log10()
```

Filter
------

```{r genes-filter}
min_count <- 1
min_cells <- 2

keep <- Matrix::rowSums(counts(filtered) >= min_count) >= min_cells

set.seed(1)
filtered <- filtered[keep, ]
sizeFactors(filtered) <- librarySizeFactors(filtered)
filtered <- normalize(filtered)
filtered <- runPCA(filtered, method = "irlba")
filtered <- runTSNE(filtered)
filtered <- runUMAP(filtered)
```

We use a minimal filter that keeps genes with at least `r min_count` counts in
at least `r min_cells` cells. After filtering we have reduced the number of
features from `r nrow(selected)` to `r nrow(filtered)`.

Validation
==========

The final quality control step is to inspect some validation plots that should
help us see if we need to make any adjustments.

Kept vs lost
------------

One thing we can look at is the difference in expression between the kept and
removed cells. If we see known genes that are highly expressed in the removed
cells that can indicate that we have removed an interesting population of cells
from the dataset. The red line shows equal expression and the blue line is a
linear fit.

```{r kept-lost}
pass_qc <- colnames(selected) %in% colnames(filtered)

lost_counts <- counts(selected)[, !pass_qc]
kept_counts <- counts(selected)[,  pass_qc]

kept_lost <- tibble(
    Gene     = rownames(selected),
    Lost     = calcAverage(lost_counts),
    LostProp = Matrix::rowSums(lost_counts > 0) / ncol(lost_counts),
    Kept     = calcAverage(kept_counts),
    KeptProp = Matrix::rowSums(kept_counts > 0) / ncol(kept_counts)
) %>%
    mutate(LogFC = predFC(cbind(Lost, Kept),
                          design = cbind(1, c(1, 0)))[, 2]) %>%
    mutate(LostCapped = pmax(Lost, min(Lost[Lost > 0]) * 0.5),
           KeptCapped = pmax(Kept, min(Kept[Kept > 0]) * 0.5))

ggplot(kept_lost,
       aes(x = LostCapped, y = KeptCapped, colour = LostProp - KeptProp)) +
    geom_point(size = 1, alpha = 0.2) +
    geom_abline(intercept = 0, slope = 1, colour = "red") +
    geom_smooth(method = "lm") +
    scale_x_log10() +
    scale_y_log10() +
    scale_colour_viridis_c() +
    xlab("Average count (removed)") +
    ylab("Average count (kept)") +
    theme_minimal()

kept_lost %>%
    select(Gene, LogFC, Lost, LostProp, Kept, KeptProp) %>%
    arrange(-LogFC) %>%
    as.data.frame() %>%
    head(100)
```

Dimensionality reduction {.tabset}
------------------------

Dimsionality reduction plots coloured by technical factors again gives us
a good overview of the dataset.

### PCA

```{r pca-valid}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotPCA(filtered, colour_by = dimred_factors[fct_name],
            shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### t-SNE

```{r t-sne-valid}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotTSNE(filtered, colour_by = dimred_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

### UMAP

```{r umap-valid}
plot_list <- lapply(names(dimred_factors), function(fct_name) {
    plotUMAP(filtered, colour_by = dimred_factors[fct_name],
             shape_by = "Sample", add_ticks = FALSE) +
        ggtitle(fct_name)
})

plot_grid(plotlist = plot_list, ncol = 2)
```

Figures
=======

Thresholds
----------

```{r fig-thresholds}
plot_data <- cell_data %>%
    mutate(Kept = !(counts_out | features_out))

exp_plot <- ggplot(plot_data,
       aes(x = log10_total_counts, y = log10_total_features_by_counts,
           colour = Kept)) +
    geom_point(alpha = 0.3) +
    geom_vline(xintercept = counts_thresh, linetype = "dashed",
               colour = "#8DC63F") +
    annotate("text", x = counts_thresh, y = Inf,
             label = paste(counts_mads, "MADs = ", round(counts_thresh, 2)),
             angle = 90, vjust = -1, hjust = 1, colour = "#8DC63F") +
    annotate("text", x = counts_thresh, y = Inf,
             label = paste(sum(counts_out), "removed"),
             angle = 90, vjust = 2, hjust = 1, colour = "#8DC63F") +
    geom_hline(yintercept = features_thresh, linetype = "dashed",
               colour = "#7A52C7") +
    annotate("text", x = Inf, y = features_thresh,
             label = paste(features_mads, "MADs = ", round(features_thresh, 2)),
             vjust = -1, hjust = 1, colour = "#7A52C7") +
    annotate("text", x = Inf, y = features_thresh,
             label = paste(sum(features_out), "removed"),
             vjust = 2, hjust = 1, colour = "#7A52C7") +
    annotate("text", x = 4.5, y = 1,
             label = paste(sum(!(counts_out | features_out)), "kept"),
             vjust = -1, colour = "#00ADEF") +
    annotate("text", x = 4.5, y = 1,
             label = paste(sum((counts_out | features_out)), "removed"),
             vjust = 1, colour = "#EC008C") +
    scale_colour_manual(values = c("#EC008C", "#00ADEF"),
                        labels = c("Removed", "Kept")) +
    ggtitle("Counts thresholds") +
    xlab(expression("Total counts ("*log["10"]*")")) +
    ylab(expression("Total features ("*log["10"]*")")) +
    theme_minimal() +
    theme(legend.position = "bottom",
          legend.title = element_blank())

mt_plot <- ggplot(cell_data, aes(x = pct_counts_MT)) +
    geom_histogram(bins = 100, fill = "#7A52C7") +
    geom_vline(xintercept = mt_thresh, linetype = "dashed",
               colour = "#EC008C") +
    annotate("text", x = mt_thresh, y = Inf,
             label = paste(mt_mads, "MADs = ", round(mt_thresh, 2), "%"),
             angle = 90, vjust = -1, hjust = 1, colour = "#EC008C") +
    annotate("text", x = mt_thresh, y = Inf,
             label = paste(sum(mt_out), "removed"),
             angle = 90, vjust = 2, hjust = 1, colour = "#EC008C") +
    ggtitle("Mitochondrial percentage threshold") +
    xlab("% counts mitochondrial") +
    ylab("Number of cells") +
    theme_minimal()

plot_data <- data.frame(Score = doublet_scores)

doublet_plot <- ggplot(plot_data, aes(x = Score)) +
    geom_histogram(bins = 100, fill = "#7A52C7") +
    geom_vline(xintercept = doublet_thresh, linetype = "dashed",
               colour = "#EC008C") +
    annotate("text", x = doublet_thresh, y = Inf,
             label = paste("Threshold =", round(doublet_thresh, 2)),
             angle = 90, vjust = -1, hjust = 1, colour = "#EC008C") +
    annotate("text", x = doublet_thresh, y = Inf,
             label = paste(sum(doublet_out), "removed"),
             angle = 90, vjust = 2, hjust = 1, colour = "#EC008C") +
    ggtitle("Doublet threshold") +
    xlab("Doublet score") +
    ylab("Number of cells") +
    theme_minimal()

plot_data <- reducedDim(selected, "PCA_coldata") %>%
    as.data.frame() %>%
    mutate(Removed = "Kept") %>%
    mutate(Removed = if_else(!kept_manual, "Manual", Removed),
           Removed = if_else(!kept_pca, "PCA", Removed),
           Removed = if_else(!kept_manual & !kept_pca, "Both", Removed)) %>%
    mutate(Removed = factor(Removed, levels = c("Both", "Manual", "Kept"),
                            labels = c("Both", "Manual only", "Kept")))

pca_plot <- ggplot(plot_data, aes(x = PC1, y = PC2, colour = Removed)) +
    geom_point(alpha = 0.3) +
    scale_colour_manual(values = c("#F47920", "#EC008C", "grey50"),
                        name = "Removed by") +
    ggtitle("PCA based cell filtering") +
    theme_minimal() +
    theme(legend.position = "bottom")

fig <- plot_grid(exp_plot, mt_plot, doublet_plot, pca_plot,
                 ncol = 2, labels = "AUTO")

ggsave(here::here("output", DOCNAME, "qc-thresholds.pdf"), fig,
       width = 7, height = 6, scale = 1.5)
ggsave(here::here("output", DOCNAME, "qc-thresholds.png"), fig,
       width = 7, height = 6, scale = 1.5)

fig
```

Validation
----------

```{r fig-validation}
plot_data <- kept_lost %>%
    mutate(Mean = 0.5 * (Lost + Kept),
           Diff = Kept - Lost)

fc_plot <- ggplot(plot_data,
                  aes(x = Mean, y  = LogFC, colour = LostProp - KeptProp)) +
    geom_point(alpha = 0.3) +
    geom_hline(yintercept = 0, colour = "red") +
    scale_x_log10(labels = scales::number_format(accuracy = 0.0001)) +
    scale_colour_viridis_c(
        name = "Prop expressed (Removed) - Prop expressed (Kept)"
    ) +
    ggtitle("Foldchange between removed and kept cells") +
    xlab("0.5 * (Average expression (Kept) + Average expression (Removed))") +
    ylab("Predicted log2 foldchange (Removed / Kept)") +
    theme_minimal() +
    theme(legend.position = "bottom")

fc_table <- kept_lost %>%
    select(Gene, `Predicted foldchange (log2)` = LogFC) %>%
    arrange(-`Predicted foldchange (log2)`) %>%
    as.data.frame() %>%
    head(15) %>%
    tableGrob(rows = NULL, theme = ttheme_minimal())

plot_data <- reducedDim(filtered, "TSNE") %>%
    as.data.frame() %>%
    rename(TSNE1 = V1, TSNE2 = V2) %>%
    mutate(
        Sample = colData(filtered)$Sample,
        `Total counts (log10)` = colData(filtered)$log10_total_counts,
        `Cell cycle` = colData(filtered)$CellCycle,
        `Total features (log10)` = colData(filtered)$log10_total_features_by_counts,
        `% Mitochondrial` = colData(filtered)$pct_counts_MT,
        `Selection method` = colData(filtered)$SelMethod
    )

plot_list <- list(
    ggplot(plot_data, aes(x = TSNE1, y = TSNE2, colour = Sample)) +
        scale_colour_manual(values = c("#EC008C", "#00ADEF", "#8DC63F")) +
        ggtitle("Sample"),
    ggplot(plot_data, aes(x = TSNE1, y = TSNE2, colour = `Selection method`)) +
        scale_colour_manual(values = c("#EC008C", "#00ADEF", "#8DC63F")) +
        ggtitle("Selection method"),
    ggplot(plot_data, aes(x = TSNE1, y = TSNE2, colour = `Cell cycle`)) +
        scale_colour_manual(values = c("#EC008C", "#00ADEF", "#8DC63F")) +
        ggtitle("Cell cycle phase"),
    ggplot(plot_data, aes(x = TSNE1, y = TSNE2, colour = `Total counts (log10)`)) +
        scale_colour_viridis_c() +
        ggtitle(expression("Total counts ("*log["10"]*")")) +
        guides(colour = guide_colourbar(barwidth = 10)),
    ggplot(plot_data,
           aes(x = TSNE1, y = TSNE2, colour = `Total features (log10)`)) +
        scale_colour_viridis_c() +
        ggtitle(expression("Total features ("*log["10"]*")")) +
        guides(colour = guide_colourbar(barwidth = 10)),
    ggplot(plot_data, aes(x = TSNE1, y = TSNE2, colour = `% Mitochondrial`)) +
        scale_colour_viridis_c() +
        ggtitle("% Mitochondrial") +
        guides(colour = guide_colourbar(barwidth = 10))
) %>%
    lapply(function(x) {
        x + geom_point(alpha = 0.3) +
            theme_minimal() +
            theme(legend.position = "bottom",
                  legend.title = element_blank())
    })

tsne_plot <- plot_grid(plotlist = plot_list, ncol = 3)

p1 <- plot_grid(fc_plot, fc_table, ncol = 2, labels = "AUTO")

fig <- plot_grid(p1, tsne_plot, ncol = 1, labels = c("", "C"),
                 rel_heights = c(0.8, 1))

ggsave(here::here("output", DOCNAME, "qc-validation.pdf"), fig,
       width = 7, height = 10, scale = 1.5)
ggsave(here::here("output", DOCNAME, "qc-validation.png"), fig,
       width = 7, height = 10, scale = 1.5)

fig
```

Summary
=======

After quality control we have a dataset with `r ncol(filtered)` cells and
`r nrow(filtered)` genes.

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters}
params <- list(
    list(
        Parameter = "counts_thresh",
        Value = counts_thresh,
        Description = "Minimum threshold for (log10) total counts"
    ),
    list(
        Parameter = "features_thresh",
        Value = features_thresh,
        Description = "Minimum threshold for (log10) total features"
    ),
    list(
        Parameter = "mt_thresh",
        Value = mt_thresh,
        Description = "Maximum threshold for percentage counts mitochondrial"
    ),
    list(
        Parameter = "doublet_thresh",
        Value = doublet_thresh,
        Description = "Maximum threshold for doublet score"
    ),
    list(
        Parameter = "counts_mads",
        Value = counts_mads,
        Description = "MADs for (log10) total counts threshold"
    ),
    list(
        Parameter = "features_mads",
        Value = features_mads,
        Description = "MADs for (log10) total features threshold"
    ),
    list(
        Parameter = "mt_mads",
        Value = features_mads,
        Description = "MADs for percentage counts mitochondrial threshold"
    ),
    list(
        Parameter = "min_count",
        Value = min_count,
        Description = "Minimum count per cell for gene filtering"
    ),
    list(
        Parameter = "min_cells",
        Value = min_cells,
        Description = "Minimum cells with min_count counts for gene filtering"
    ),
    list(
        Parameter = "n_cells",
        Value = ncol(filtered),
        Description = "Number of cells in the filtered dataset"
    ),
    list(
        Parameter = "n_genes",
        Value = nrow(filtered),
        Description = "Number of genes in the filtered dataset"
    ),
    list(
        Parameter = "median_genes",
        Value = median(Matrix::colSums(counts(filtered) != 0)),
        Description = paste("Median number of expressed genes per cell in the",
                            "filtered dataset")
    ),
    list(
        Parameter = "median_counts",
        Value = median(Matrix::colSums(counts(filtered))),
        Description = paste("Median number of counts per cell in the filtered",
                            "dataset")
    )
)


params <- jsonlite::toJSON(params, pretty = TRUE)
knitr::kable(jsonlite::fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r save}
write_rds(filtered, here::here("data/processed/02-filtered.Rds"))
```

```{r output}
dir.create(here::here("output", DOCNAME), showWarnings = FALSE)

readr::write_lines(params, here::here("output", DOCNAME, "parameters.json"))

knitr::kable(data.frame(
    File = c(
        getDownloadLink("parameters.json", DOCNAME),
        getDownloadLink("qc-thresholds.png", DOCNAME),
        getDownloadLink("qc-thresholds.pdf", DOCNAME),
        getDownloadLink("qc-validation.png", DOCNAME),
        getDownloadLink("qc-validation.pdf", DOCNAME)
    ),
    Description = c(
        "Parameters set and used in this analysis",
        "Quality control thresholds figure (PNG)",
        "Quality control thresholds figure (PDF)",
        "Quality control validation figure (PNG)",
        "Quality control validation figure (PDF)"
    )
))
```

Session information
-------------------

```{r session-info, cache = FALSE}
devtools::session_info()
```
