---
title: "PAGA"
---

```{r knitr, include = FALSE}
DOCNAME = "05-paga"
NOW <- Sys.time()

# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
    if (before) {
        print(paste("Start:", Sys.time()))
        NOW <<- Sys.time()
    } else {
        print(paste("Stop:", Sys.time()))
        print(Sys.time() - NOW)
    }
})

knitr::opts_chunk$set(
    autodep        = TRUE,
    cache          = TRUE,
    cache.path     = paste0("cache/", DOCNAME, "/"),
    cache.comments = FALSE,
    echo           = TRUE,
    error          = FALSE,
    fig.align      = "center",
    fig.width      = 10,
    fig.height     = 8,
    message        = FALSE,
    warning        = FALSE,
    timeit         = TRUE
)
```

```{r libaries, cache = FALSE}
# scRNA-seq
library("SingleCellExperiment")

# System
library("processx")

# Plotting
library("cowplot")

# Presentation
library("knitr")

# Tidyverse
library("tidyverse")
```

```{r source, cache = FALSE}
source(here::here("R/output.R"))
source(here::here("R/plotting.R"))
```

```{r depends-paths}
clust_path <- here::here("data/processed/03-clustered.Rds")
paga_path <- here::here("scripts/paga.py")
```

```{r bpparam, cache = FALSE}

```

Introduction
============

In this document we are going to explore the relationship between our clusters
using partition-based graph abstraction (PAGA). PAGA builds a graph where each
node is a cluster and the edges represent connectivity between clusters. This
can give us an overview of the dataset and the relationships between cell types.

```{r load-sce, cache.extra = tools::md5sum(clust_path)}
if (file.exists(clust_path)) {
    sce <- read_rds(clust_path)
} else {
    stop("Clustered dataset is missing. ",
         "Please run '03-clustering.Rmd' first.",
         call. = FALSE)
}
```

PAGA
====

PAGA is available as part of the `scanpy` Python package. The PAGA analysis is
performed using the following Python script.

```{r paga, comment = "", cache.extra = c(tools::md5sum(paga_path), tools::md5sum(clust_path))}
cat(readLines(paga_path), sep = "\n")

paga_out <- run(here::here("scripts/run_paga.sh"))
```

We then load the PAGA results and visualise them here.

```{r load-paga}
clust_embedding <- read_csv(
    here::here("output/05-paga/cluster_embedding.csv"),
    col_types = cols(
        .default = col_double()
    )
) %>%
    mutate(Size = as.numeric(table(colData(sce)$Cluster))) %>%
    mutate(Cluster = factor(Cluster, levels = levels(colData(sce)$Cluster)))

clust_edges <- read_csv(
    here::here("output/05-paga/cluster_edges.csv"),
    col_types = cols(
        .default = col_double()
    )
) %>%
    mutate(To = factor(To, levels = levels(colData(sce)$Cluster)),
           From = factor(From, levels = levels(colData(sce)$Cluster))) %>%
    left_join(clust_embedding, by = c("From" = "Cluster")) %>%
    rename(FromX = X, FromY = Y) %>%
    select(-Size) %>%
    left_join(clust_embedding, by = c("To" = "Cluster")) %>%
    rename(ToX = X, ToY = Y) %>%
    select(-Size)

cell_embedding <- read_csv(
    here::here("output/05-paga/cell_embedding.csv"),
    col_types = cols(
        .default = col_double(),
        Cell = col_character()
    )
) %>%
    mutate(Cluster = colData(sce)$Cluster)

cell_edges <- read_csv(
    here::here("output/05-paga/cell_edges.csv"),
    col_types = cols(
        .default = col_double(),
        From = col_character(),
        To = col_character()
    )
) %>%
    left_join(cell_embedding, by = c("From" = "Cell")) %>%
    rename(FromX = X, FromY = Y) %>%
    select(-Cluster) %>%
    left_join(cell_embedding, by = c("To" = "Cell")) %>%
    rename(ToX = X, ToY = Y) %>%
    select(-Cluster)
```

Cluster graph
=============

The PAGA graph shows the relationship between clusters. Here we have use a
force directed graph layout in order to visualise it. PAGA calculates the
connectivity between each pair of clusters but we need to apply some threshold
to that to select the meaningful edges.

Thresholds {.tabset}
----------

```{r clust-paga, results = "hide"}
src_list <- lapply(seq(0, 0.9, 0.1), function(thresh) {
    src <- c(
        "### Con {{thresh}} {.unnumbered}",
        "```{r clust-paga-{{thresh}}}",
        "plotPAGAClustGraph(clust_embedding, clust_edges, thresh = {{thresh}})",  
        "```",
        ""
    )
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Edges by threshold
------------------

Number of selected edges for different threshold connectivities.

```{r thresh-edges}
plot_data <- tibble(
    Threshold = seq(0, 1, 0.01)
) %>%
    mutate(Edges = map_int(Threshold, function(thresh) {
        sum(clust_edges$Connectivity > thresh)
    }))

con_thresh <- 0.7

ggplot(plot_data, aes(x = Threshold, y = Edges)) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept = con_thresh, colour = "red") +
    xlab("Connectivity threshold") +
    ylab("Number of edges") +
    theme_minimal()
```

Here we have selected a connectivity threshold of `r con_thresh`.

Cell graph
==========

The cluster graph can be used as a starting point to layout individual cells,
allowing us to see things at a higher resolution.

```{r cell-paga}
plotPAGACellGraph(cell_embedding, cell_edges, thresh = 0.1)
```

Compare
=======

Looking the two views together let's us see both global and specific details.

```{r compare-paga}
plotPAGACompare(clust_embedding, clust_edges, clust_thresh = con_thresh,
                cell_embedding, cell_edges, cell_thresh = 0.1)
```

Genes {.tabset}
-----

```{r compare-genes, results = "hide"}
known_genes <- c(
    # Stroma
    "TAGLN", "ACTA2", "MAB21L2", "DLK1", "GATA3", "COL2A1", "COL9A3",
    # Podocyte
    "PODXL", "NPHS2", "TCF21",
    # Cell cycle
    "HIST1H4C", "PCLAF", "CENPF", "HMGB2",
    # Endothelium
    "CLDN5", "PECAM1", "KDR", "CALM1",
    # Neural
    "TTYH1", "SOX2", "HES6", "STMN2",
    # Epithelium
    "PAX2", "PAX8", "KRT19",
    # Muscle
    "MYOG", "MYOD1"
)

for (gene in known_genes) {
    cell_embedding[[gene]] <- logcounts(sce)[gene, ]
}

clust_genes <- cell_embedding %>%
    select(-Cell, -X, -Y) %>%
    group_by(Cluster) %>%
    summarise_all(mean)

clust_embedding <- left_join(clust_embedding, clust_genes, by = "Cluster")

src_list <- lapply(known_genes, function(gene) {
    src <- c(
        "### {{gene}} {.unnumbered}",
        "```{r compare-{{gene}}}",
        "plotPAGACompare(clust_embedding, clust_edges,",
        "clust_thresh = con_thresh, cell_embedding, cell_edges,
        cell_thresh = 0.1, colour = '{{gene}}')",  
        "```",
        ""
    )
    knit_expand(text = src)
})

out <- knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

Figures
=======

```{r figure}
clust_plot <- ggplot(clust_embedding, aes(x = X, y = Y)) +
    geom_segment(data = clust_edges,
                 aes(x = FromX, y = FromY, xend = ToX, yend = ToY,
                     colour = Connectivity, alpha = Connectivity), size = 4) +
    scale_colour_viridis_c(limits = c(0, 1)) +
    scale_alpha_continuous(limits = c(0, 1), range = c(0, 1), guide = FALSE) +
    geom_point(aes(fill = Cluster, size = Size), shape = 21) +
    geom_text(aes(label = Cluster)) +
    scale_size(range = c(6, 20), guide = FALSE) +
    scale_fill_discrete(guide = FALSE) +
    guides(colour = guide_colourbar(barwidth = 20)) +
    ggtitle("PAGA cluster graph") +
    theme_void() +
    theme(plot.title = element_text(size = rel(1.2), hjust = 0.1, 
                                    vjust = 1, margin = margin(5)),
          legend.position = "bottom")

cell_plot <- ggplot(cell_embedding, aes(x = X, y = Y)) +
    geom_segment(data = cell_edges,
                 aes(x = FromX, y = FromY, xend = ToX, yend = ToY,
                     size = Connectivity, alpha = Connectivity)) +
    geom_point(aes(colour = Cluster), size = 1) +
    scale_alpha_continuous(limits = c(0, 1), range = c(0, 0.5), guide = FALSE) +
    scale_colour_discrete(guide = guide_legend(
        nrow = 1, override.aes = list(size = 4)
    )) +
    scale_size(range = c(0.1, 0.5), guide = FALSE) +
    ggtitle("Cell shared nearest neighbour graph") +
    theme_void() +
    theme(plot.title = element_text(size = rel(1.2), hjust = 0.1, 
                                    vjust = 1, margin = margin(5)),
          legend.position = "bottom")

fig <- plot_grid(clust_plot, cell_plot, nrow = 2, labels = "AUTO")

ggsave(here::here("output", DOCNAME, "paga-results.pdf"), fig,
       width = 7, height = 8, scale = 1.5)
ggsave(here::here("output", DOCNAME, "paga-results.png"), fig,
       width = 7, height = 8, scale = 1.5)

fig
```

Summary
=======

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters, cache.lazy = FALSE}
params <- list(
    list(
        Parameter = "con_thresh",
        Value = con_thresh,
        Description = "Connectivity threshold for PAGA graph"
    )
)

params <- jsonlite::toJSON(params, pretty = TRUE)
knitr::kable(jsonlite::fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r save}

```

```{r output}
dir.create(here::here("output", DOCNAME), showWarnings = FALSE)

readr::write_lines(params, here::here("output", DOCNAME, "parameters.json"))

knitr::kable(data.frame(
    File = c(
        getDownloadLink("parameters.json", DOCNAME),
        getDownloadLink("cluster_embedding.csv", DOCNAME),
        getDownloadLink("cluster_edges.csv", DOCNAME),
        getDownloadLink("cluster_tree_edges.csv", DOCNAME),
        getDownloadLink("cell_embedding.csv", DOCNAME),
        getDownloadLink("cell_edges.csv", DOCNAME),
        getDownloadLink("paga-results.png", DOCNAME),
        getDownloadLink("paga-results.pdf", DOCNAME)
    ),
    Description = c(
        "Parameters set and used in this analysis",
        "Embedding for clusters from PAGA",
        "Edges for PAGA cluster graph",
        "Tree edges for PAGA cluster graph",
        "Embedding for cells from PAGA",
        "Edges for cell graph",
        "PAGA results figure (PNG)",
        "PAGA results figure (PDF)"
    )
))
```

Session information
-------------------

```{r session-info, cache = FALSE}
devtools::session_info()
```
