---
title: "Velocyto"
---

```{r knitr, include = FALSE}
DOCNAME = "06-velocyto"
NOW <- Sys.time()

# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
    if (before) {
        print(paste("Start:", Sys.time()))
        NOW <<- Sys.time()
    } else {
        print(paste("Stop:", Sys.time()))
        print(Sys.time() - NOW)
    }
})

knitr::opts_chunk$set(
    autodep        = TRUE,
    cache          = TRUE,
    cache.path     = paste0("cache/", DOCNAME, "/"),
    cache.comments = FALSE,
    echo           = TRUE,
    error          = FALSE,
    fig.align      = "center",
    fig.width      = 10,
    fig.height     = 8,
    message        = FALSE,
    warning        = FALSE,
    timeit         = TRUE
)
```

```{r libaries, cache = FALSE}
# scRNA-seq
library("SingleCellExperiment")
library("velocyto.R")

# Tidyverse
library("tidyverse")
```

```{r source, cache = FALSE}
source(here::here("R/output.R"))
```

```{r depends-paths}
clust_path <- here::here("data/processed/03-clustered.Rds")
loom1_path <- here::here("data/velocyto/Org1.loom")
loom2_path <- here::here("data/velocyto/Org2.loom")
loom3_path <- here::here("data/velocyto/Org3.loom")
loom_paths <- c(loom1_path, loom2_path, loom3_path)
```

```{r bpparam, cache = FALSE}

```

Introduction
============

In this document we are to perform cell velocity analysis using velocyto. This
approach look at the number of spliced and unspliced reads from each gene and
attempts to identify which are being actively transcribed and therefore the
direction each cell is differentiating towards.

```{r load-sce, cache.extra = tools::md5sum(clust_path)}
if (file.exists(clust_path)) {
    sce <- read_rds(clust_path)
} else {
    stop("Clustered dataset is missing. ",
         "Please run '03-clustering.Rmd' first.",
         call. = FALSE)
}

warning("New clustered dataset loaded, check Loom files are up to date!")
```

The first step in using velocyto is to process the aligned BAM files to
separately count spliced and unspliced reads. This is done using a command line
program and can take a relatively long time so here we start by reading in
those results.

```{r load-velocyto, cache.extra = tools::md5sum(loom_paths), results = "hide"}
org1 <- read.loom.matrices(loom1_path)
org2 <- read.loom.matrices(loom2_path)
org3 <- read.loom.matrices(loom3_path)

spliced <- cbind(org1$spliced, org2$spliced, org3$spliced)
unspliced <- cbind(org1$unspliced, org2$unspliced, org3$unspliced)

cell_idx <- colData(sce)$Cell
names(cell_idx) <- paste0(
    "Org", colData(sce)$Sample, ":", colData(sce)$Barcode, "x"
)

colnames(spliced) <- unname(cell_idx[colnames(spliced)])
spliced <- spliced[intersect(rownames(sce), rownames(spliced)), colnames(sce)]

colnames(unspliced) <- unname(cell_idx[colnames(unspliced)])
unspliced <- unspliced[intersect(rownames(sce), rownames(unspliced)),
                       colnames(sce)]
```

Velocyto
========

We now run velocyto to calclate velocities for each cell. We also project these
results onto our previous dimensionality reductions for visualisation.

```{r velocyto, results = "hide", fig.show = "hide"}
spliced <- filter.genes.by.cluster.expression(
    spliced, colData(sce)$Cluster,
    min.max.cluster.average = 0.2
)

unspliced <- filter.genes.by.cluster.expression(
    unspliced, colData(sce)$Cluster,
    min.max.cluster.average = 0.05
)

velocity <- gene.relative.velocity.estimates(
    spliced, unspliced,
    deltaT       = 1,
    kCells       = 30,
    fit.quantile = 0.02,
    n.cores      = 10
)

vel_embedding <- show.velocity.on.embedding.cor(
    reducedDim(sce, "SeuratTSNE"), velocity,
    n.cores = 10, show.grid.flow = TRUE, return.details = TRUE
)
```

By cell
-------

This plot shows the velocity of individual cells. The direction of each arrow
indicates where each cell is headed in this space based on the genes that are
being actively transcribed and the length is an indication of rate.

```{r velo-tsne}
plot_data <- reducedDim(sce, "SeuratTSNE") %>%
    as.data.frame() %>%
    mutate(X0 = vel_embedding$arrows[, "x0"],
           X1 = vel_embedding$arrows[, "x1"],
           Y0 = vel_embedding$arrows[, "y0"],
           Y1 = vel_embedding$arrows[, "y1"]) %>%
    mutate(X2 = X0 + (X1 - X0) * 4,
           Y2 = Y0 + (Y1 - Y0) * 4) %>%
    mutate(Cluster = colData(sce)$Cluster)

ggplot(plot_data) +
    geom_point(aes(x = tSNE_1, y = tSNE_2, colour = Cluster)) +
    geom_segment(aes(x = X0, xend = X2, y = Y0, yend = Y2),
                 arrow = arrow(length = unit(3, "points"), type = "closed"),
                 colour = "grey20", alpha = 0.8) +
    theme_minimal()
```

Vector field
------------

It can be hard to see all the arrows for individual cells and noise in the data
can make similar cells appear to be heading in different directions. Here we
summarise the data by building a grid field of vectors that show the average
velocity of nearby cells. This gives us a global view of the transcriptional
direction of of the dataset.

```{r velo-tsne-field}
field_arrows <- vel_embedding$garrows %>%
    as.data.frame() %>%
    mutate(x2 = x0 + (x1 - x0) * 10,
           y2 = y0 + (y1 - y0) * 10)

ggplot(plot_data) +
    geom_point(aes(x = tSNE_1, y = tSNE_2, colour = Cluster)) +
    geom_segment(data = field_arrows,
                 aes(x = x0, xend = x2, y = y0, yend = y2),
                 size = 1,
                 arrow = arrow(length = unit(4, "points"), type = "closed"),
                 colour = "grey20", alpha = 0.8) +
    theme_minimal()
```

Summary
=======

Parameters
----------

This table describes parameters used and set in this document.

```{r parameters, cache.lazy = FALSE}
params <- list(

)

names(params) <- map_chr(params, magrittr::extract2, "Parameter")
metadata(sce)$Params[[DOCNAME]] <- params

names(params) <- NULL
params <- jsonlite::toJSON(params, pretty = TRUE)
knitr::kable(jsonlite::fromJSON(params))
```

Output files
------------

This table describes the output files produced by this document. Right click
and _Save Link As..._ to download the results.

```{r save}

```

```{r output}
dir.create(here::here("output", DOCNAME), showWarnings = FALSE)

knitr::kable(data.frame(
    File = c(
        getDownloadLink("parameters.json", DOCNAME)
    ),
    Description = c(
        "Parameters set and used in this analysis"
    )
))
```

Session information
-------------------

```{r session-info, cache = FALSE}
devtools::session_info()
```
